Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-18 paper/manuscript.tex
Synchronise changes made in Overleaf, incuding:
- Undated discussion of uncertainty estimation by Agastya Bhati and Peter Coveney (changes still required to address review)
- ATM section with comments from Darrin York addressed.
- Updated discussion of non-equilibrium switching by Vytas Gapsys.
- Updates to the "importance of conformational sampling" section made by Wei-Tse Hsu in response to comments from Emilio Gallicchio.
- Various other minor tweaks

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[11]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}

%
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}

@@

Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states, but these are not our focus here.~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}

@@

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e.with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial velocities. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat run as well as different velocities.

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. Further work is needed to ensure reproducibility of alchemical free energy calculations across different software implementations to guarantee that force-field development efforts lead to transferable potential energy functions.

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This can happen in buried binding sites~\cite{laage2017water}. Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations. In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation. If the locations of the bound water molecules differs at the two endpoints - e.g. if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation. Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.

In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state. In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse). When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue. However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.

\paragraph{Membrane proteins require additional care}
Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations. Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}). Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}. These systems require more extensive equilibration than soluble proteins. Some workflows, such as CHARMM-GUI or Desmond membrane preparation, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.

Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water. Several approaches are reviewed in \cite{wu2022correction-51c}. The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.

In the case of a lipid-facing binding site, there are three additional challenges. First, simulation time should be extended to ensure sufficient sampling of lipid configurational space. Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand. If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols. Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids. Second, considering the free energy cost of a ligand partitioning into the membrane may provide additional insights \cite{doi:10.1021/acs.jcim.1c01147}. Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species. In some cases, the identity of the lipid may be known. But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.

@@

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

@@

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. & 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors. \cite{ries2024Kartograf}

@@

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}

@@

The need for and role of dummy atoms is clearly visible in Fig.~\ref{fig:fig_topology}: One or more dummy atoms are attached to the physical molecule. A crucial requirement is that these atoms, typically connected through bonded energy terms (bond stretching, angle bending, dihedral angles) do not affect equilibrium and dynamical properties of the physical system. However, since a physical molecule + dummy atoms attached to it can or must be viewed as a modified physical molecule, the partition function of the molecules with and without dummy atoms will NOT be identical. In other words, a single free energy difference between two molecules A and B will be different from that of A(+D) and B(+D), where the (+D) indicates the presence of one or more dummy atoms needed so that the alchemical transformation can be carried out. When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function. By construction, relative free energy simulatins involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs). If dummy atoms are interacting with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function. As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} aoid this problem.

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[11]{Agastya P. Bhati}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[11, 12, 13]{Peter V. Coveney}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
%
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
\orcid{Peter V. Coveney}{0000-0002-8787-7256}

@@

Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states (see Sec.~\ref{sec:sampling_schemes}).~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}

@@

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e.with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. 
There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions. Coveney \textit{et al. }~\cite{coveney2016} have noted that, for a complex dynamical system to reach equilibrium, the dynamics must at least be ``mixing'' which is an established result in ergodic theory. A system which is ``mixing'' is also ``ergodic'' by definition but the reverse is not true. This implies that MD trajectories are chaotic, in other words they are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible. Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}. In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e. statistically robust) free energy predictions. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017} which is based on the use of ensemble methods. It produces excellent results for large scale benchmarking datasets for both RBFEs~\cite{bhati2022} and ABFEs~\cite{bhati2025}. It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision. Further, this method is extremely flexible and adaptive. For example, one can simply add additional ensemble members as necessary or desired, without having to re-run anything done initially. It should also be noted that ensemble simulations are computationally more effective than long simulations. On a large supercomputer, all the sampling can be done concurrently obtaining the statistical results in the time it takes to run a single simulation. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each replica as well as different velocities.

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. 
A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}. The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties. To conclude, ensemble simulations can remove the aleatoric uncertainties in MD based predictions, however, further work is needed to ensure reproducibility across software packages arising due to systematic/parametric uncertainties.

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}. Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points. This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points. The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points. The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.

@@

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023,York_ACSPhysChemAu_2023_v3_p478}.
While the implementation technically uses two sets of coordinates for the entire transforming region, selected atoms that have a direct atom-atom mapping are constrained to have common coordinates (sometimes referred to as a "common core"), while others that are mapped into dummy atoms and are treated with softcore potentials~\cite{Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640} have separable coordinates.  Thus the common core part of the dual topology are constrained to have the same coordinates to maximize phase space overlap, whereas the softcore region has separate coordinates.  The two topologies do not see each other, and this property enables this approach to be leveraged in the alchemical enhanced sampling (ACES) method~\cite{Lee_2023,Tsai_JChemInfModel_2024_v64_p7046,Zhang_JChemTheoryComput_2024_v20_p3935}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

@@

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. & 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors. \cite{ries2024Kartograf}

@@

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, Rieder2022Lev}

@@

The need for and role of dummy atoms is clearly visible in Fig.~\ref{fig:fig_topology}: One or more dummy atoms are attached to the physical molecule. A crucial requirement is that these atoms, typically connected through bonded energy terms (bond stretching, angle bending, dihedral angles) do not affect equilibrium and dynamical properties of the physical system. However, since a physical molecule + dummy atoms attached to it can or must be viewed as a modified physical molecule, the partition function of the molecules with and without dummy atoms will NOT be identical. In other words, a single free energy difference between two molecules A and B will be different from that of A(+D) and B(+D), where the (+D) indicates the presence of one or more dummy atoms needed so that the alchemical transformation can be carried out. When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function. By construction, relative free energy simulations involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs). If dummy atoms  interact with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function. As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, TIES~\cite{bhati2017, bieniek2021} circumvents this issue by design as it does not need any restraints.
